---
output:
  pdf_document: default
  html_document: default
---
# Trees
a) Investigate whether the tree type influences volume by performing ANOVA, without taking diameter and height into account. Can a t-test be related to the above ANOVA test? Estimate the volumes for the two tree types.

```{r}
#Load the data and calculate the mean volume for each tree type
data <- read.table("treeVolume.txt", header=TRUE)
mean_vol_beech <- mean(data[data$type == "beech", "volume"])
mean_vol_oak <- mean(data[data$type == "oak", "volume"])
```

We use the aov() function to perform ANOVA:

```{r}
aov_result <- aov(volume ~ type, data=data)
summary(aov_result)
```

The output of the summary function shows the results of the ANOVA test. When the p-value is below the given confidence level (0.05), we can conclude that there is a significant difference in volumes between the two tree types.

We can see that the p-value is 0.174. Therefore we can conclude that there is not a significant difference in volumes between the two tree types.

It is possible to relate the ANOVA test to a t-test by performing a two-sample t-test between the volumes of the two tree types. However, the ANOVA test is more appropriate when we have more than two groups to compare.

It is possible to relate an ANOVA test to a t-testby performing a two-sample t-test between the volumes of the two tree types. However, the ANOVA test is more appropriate when we have more than two groups to compare.

To perform a t-test, we can use the t.test() function:
```{r}
t_test_result <- t.test(data[data$type == "beech", "volume"], data[data$type == "oak", "volume"])
t_test_result
```

The p-value from the t-test is 0.1659. Relatively similar result to ANOVA performed earlier. The t-test shows us if there is a significant difference in mean volume between the two tree types. With the result we can again say that there is not a significant difference in volumes between the two tree types.

To calculate the mean value of each group we estimate the volumes for each of the tree types.
```{r}
# calculate mean volumes for each tree type
mean_volumes <- tapply(data$volume, data$type, mean)
mean_volumes
```

b)  Now include diameter and height as explanatory variables into the analysis. Investigate whether the influence of diameter on volume is similar for the both tree types. Do the same for the influence of height on volume. (Consider at most one (relevant) pairwise interaction per model.) Comment.

To investigate the influence of diameter and height on volume, we can perform a multiple linear regression analysis in R. We can start by creating a model that includes both diameter and height as explanatory variables:

```{r}
# create multiple linear regression model
reg_model <- lm(volume ~ type + diameter + height, data = data)
summary(reg_model)
```

This code will create a multiple linear regression model using the lm() function in R and display the results using the summary() function. The output will show the coefficients for each variable and their significance.

To investigate whether the influence of diameter on volume is similar for the both tree types, we can include a diameter-tree type interaction term in the model:
```{r}
# create multiple linear regression model with interaction term
reg_model_interaction <- lm(volume ~ type * diameter + height, data = data)
summary(reg_model_interaction)
```

This code will create a new regression model with an interaction term between tree type and diameter, and display the results using the summary() function. We can see that the interaction term is not significant (0.474 > 0.05), therefore, the relationship between diameter and volume does not depend on the tree type.

We can perform a similar analysis to investigate the influence of height on volume:
```{r}
# create multiple linear regression model with interaction term
reg_model_interaction <- lm(volume ~ type + diameter + height * type, data = data)
summary(reg_model_interaction)
```

This code will create a new regression model with an interaction term between tree type and height, and display the results using the summary() function. We can see that the interaction term is not significant (0.17613 > 0.05), therefore, the relationship between height and volume does not depend on the tree type.

In conclusion, by including diameter and height as explanatory variables, we can see that both variables do not have a significant influence on volume, and that the relationships between these variables and volume do not depend on the tree type.

c)  Using the results from c), investigate how diameter, height and type influence volume. Comment. Using the resulting model, predict the volume for a tree with the (overall) average diameter and height?

Using the results from b), we can fit a linear regression model for volume using diameter, height, and type as explanatory variables:
```{r}
lm_all <- lm(volume ~ diameter + height + type, data=data)
summary(lm_all)
```
The results show that diameter and height have a significant influence on volume. Type does not have a significant influence on volume.

To predict the volume for a tree with the overall average diameter and height, we can calculate the mean values for diameter and height and use them in the model:
```{r}
mean_diam <- mean(data$diameter)
mean_height <- mean(data$height)

predicted_vol <- predict(lm_all, newdata=data.frame(diameter=mean_diam, height=mean_height, type="beech"))
predicted_vol
```

d)  Propose a transformation of the explanatory variables that possibly yields a better model (verify this).  (Hint: think of a natural link between the response and explanatory variables.)

One possible transformation of the explanatory variables is to take the natural logarithm of the diameter and height. This transformation can be useful when the relationship between the explanatory variables and the response is not linear but instead follows a logarithmic pattern. We can apply this transformation by creating new variables in the data frame:

```{r}
data$log_diameter <- log(data$diameter)
data$log_height <- log(data$height)
```
Then, we can build a new model using these transformed variables:
```{r}
model2 <- lm(volume ~ log_diameter + log_height + type, data = data)
summary(model2)
```
We can compare the R-squared values of the two models to see which one is a better fit. If the R-squared value for model2 is higher, then the logarithmic transformation improved the model fit. If not, then the original model may be the better choice.