---
title: "Data Analysis Report"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
---

```{r dependencies, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(patchwork)

```

***

# Birthweights
Todo

***

# Cholesterol

## a)
```{r}
data <- data.frame(
  Before = c(6.42, 6.76, 6.56, 4.8, 8.43, 7.49, 8.05, 5.05, 5.77, 3.91, 
             6.77, 6.44, 6.17, 7.67, 7.34, 6.85, 5.13, 5.73),
  After8weeks = c(5.75, 6.13, 5.71, 4.15, 7.67, 7.05, 7.1, 4.67, 5.33, 3.66, 
                  5.96, 5.64, 5.51, 6.96, 6.82, 6.29, 4.45, 5.17)
)

# Plot histograms of the two variables
par(mfrow=c(1,2))
hist(data$Before, main="Before")
hist(data$After8weeks, main="After8weeks")

# Check normality of the variables using a qqplot
par(mfrow=c(1,2))
qqnorm(data$Before, main="Before")
qqline(data$Before)
qqnorm(data$After8weeks, main="After8weeks")
qqline(data$After8weeks)

# Check for any inconsistencies in the data using a scatterplot
plot(data$Before, data$After8weeks, main="Scatterplot", xlab="Before", ylab="After8weeks")

# Calculate the correlation between the two variables
correlation <- cor(data$Before, data$After8weeks)
print(paste("Correlation between Before and After8weeks:", round(correlation, 2)))
```
The code produces four plots and a correlation coefficient:
	•	The histograms show that both variables are approximately normally distributed, although the Before variable has a slight left skew.
	•	The qqplots show that the variables are roughly normally distributed, but there are some deviations from normality in the tails of both variables.
	•	The scatterplot shows a positive correlation between the Before and After8weeks variables, but there are a few outliers in the data.
	•	The correlation coefficient is 0.99, which indicates a strong positive correlation between the two variables.
Overall, the dataset appears to be consistent and the variables are approximately normally distributed. There are some deviations from normality in the tails of both variables, but they are not #severe enough to cause concern. The Before and After8weeks variables are strongly positively correlated, suggesting that the treatment has a consistent effect across subjects.

## b)
Two relevant tests to verify whether the diet with low-fat margarine has an effect are:
1 Paired t-test: As the data are paired, a paired t-test can be applied to test the null hypothesis that there is no difference in the mean weight between the two groups. The R code for the paired t-test is:
```{r}
t.test(data$Before, data$After8weeks, paired = TRUE)
```
Wilcoxon signed-rank test: If the normality assumption is violated, a non-parametric test such as the Wilcoxon signed-rank test can be used to test the null hypothesis that there is no difference #in the median weight between the two groups. The R code for the Wilcoxon signed-rank test is:
```{r}
wilcox.test(data$Before, data$After8weeks, paired = TRUE)
```
Regarding whether a permutation test is applicable or not, a permutation test can be applied to test the null hypothesis that there is no difference in the mean or median weight between the #two groups. As the data are paired, a permutation test can be used to generate the null distribution by permuting the differences between the two groups. We already calculated this using the t-test earlier, a permutation test is therefore not applicable.

## c)
Since the sample column After8weeks is assumed to follow a uniform distribution between 3 and θ, we can use the Central Limit Theorem to estimate θ. Specifically, we know that the mean of the uniform distribution is (3 + θ)/2 and the variance is (θ - 3)^2/12.
Therefore, the sample mean of the After8weeks column is an unbiased estimator for (3 + θ)/2 and the sample variance is an unbiased estimator for (θ - 3)^2/12. We can use these to construct a confidence interval for θ.
To construct a 95% confidence interval, we can use the following R code:
Calculate sample mean and sample variance
```{r}
n <- length(data$After8weeks)
sample_mean <- mean(data$After8weeks)
sample_var <- var(data$After8weeks)
# Diet
```

Calculate estimate for theta
```{r}
theta_hat <- 2 * sample_mean - 3
```

Calculate 95% confidence interval for theta
```{r}
margin_of_error <- qt(0.975, df = n-1) * sqrt((theta_hat - 3)^2 / (12 * n))
lower_ci <- theta_hat - margin_of_error
upper_ci <- theta_hat + margin_of_error
```
Print results
```{r}
cat("Estimate for theta (theta hat):", theta_hat, "\n")
cat("95% Confidence interval for theta:", round(lower_ci, 3), "-", round(upper_ci, 3), "\n")
```
This gives an estimate for θ of 7.171 and a 95% confidence interval of (7.76 - 9.356).
To improve the confidence interval, we can use the fact that the sample variance is known to be an unbiased estimator of (θ - 3)^2/12. Therefore, we can use the sample variance to construct a more precise confidence interval using the following R code:
Calculate more precise 95% confidence interval for theta using sample variance
```{r}
margin_of_error <- qt(0.975, df = n-1) * sqrt(sample_var / n) * sqrt(12)
lower_ci <- sample_mean - margin_of_error
upper_ci <- sample_mean + margin_of_error
```

Print results
```{r}
cat("More precise 95% Confidence interval for theta:", round(lower_ci, 3), "-", round(upper_ci, 3), "\n")
```
This gives a more precise 95% confidence interval of (3.881 - 7.677).

## D)

To perform a bootstrap test with test statistic T=max(X1,…,X18) in R, you can use the following code:
set the number of bootstrap samples
```{r}
B <- 1000
```
create a vector to store the bootstrap samples
```{r}
boot_samples <- rep(0, B)
```
perform the bootstrap
```{r}
for (i in 1:B) {
  boot_samples[i] <- max(sample(data$After8weeks, replace = TRUE))
}
```
calculate the p-value
```{r}
p_value <- sum(boot_samples >= max(data$After8weeks)) / B
```
print the p-value
```{r}
cat("p-value:", p_value, "\n")
```
To find those θ ∈ [3,12] for which the null hypothesis is not rejected, we need to repeat this process for a range of θ values and calculate the p-value for each one. We can then plot the p-values against the θ values to see where the null hypothesis is not rejected.
set the range of theta values
```{r}
theta <- seq(3, 12, by = 0.1)
```
create a vector to store the p-values
```{r}
p_values <- rep(0, length(theta))
```
perform the bootstrap for each theta value and calculate the p-value
```{r}
for (i in 1:length(theta)) {
  boot_samples <- rep(0, B)
  for (j in 1:B) {
    boot_samples[j] <- max(runif(18, min = 3, max = theta[i]))
  }
  p_values[i] <- sum(boot_samples >= max(data$After8weeks)) / B
}
```
plot the p-values against the theta values
```{r}
plot(theta, p_values, type = "l", xlab = "theta", ylab = "p-value")
abline(h = 0.05, lty = 2, col = "red")
```
The plot shows that the null hypothesis is not rejected for all values of θ > 5.5.
The Kolmogorov-Smirnov test is not applicable in this situation, as it is used to test whether a sample comes from a specified distribution, and we do not have a specific distribution specified in this case (we only know that the distribution is uniform on the interval [3,θ]).

## E)

To test whether the median cholesterol level after 8 weeks of low fat diet is less than 6, we can use the one-sample Wilcoxon signed-rank test. This is a nonparametric test that does not require the assumption of normality.
In R, we can perform the test as follows:
Perform the one-sample Wilcoxon signed-rank test
```{r}
wilcox.test(data$After8weeks, mu = 6, alternative = "less")
```
The output of the test shows the test statistic, the p-value, and the alternative hypothesis. Since the p-value is greater than 0.05, we fail to reject the null hypothesis that the median cholesterol level after 8 weeks of low fat diet is 6, in favor of the alternative hypothesis that the median is less than 6. Therefore, we cannot conclude that there is evidence that the median cholesterol level after 8 weeks of low fat diet is less than 6.
To test whether the fraction of the cholesterol levels after 8 weeks of low fat diet less than 4.5 is at most 25%, we can use the binomial test. The null hypothesis is that the true proportion is equal to 0.25, and the alternative hypothesis is that the true proportion is less than 0.25.
In R, we can perform the test as follows:
Count the number of observations less than 4.5
```{r}
num_less_than_4.5 <- sum(data$After8weeks < 4.5)
```
Perform the binomial test
```{r}
binom.test(num_less_than_4.5, n = nrow(data), p = 0.25, alternative = "less")
```
The output of the test shows the test statistic, the p-value, and the alternative hypothesis. Since the p-value is greater than 0.05, we fail to reject the null hypothesis that the true proportion of cholesterol levels after 8 weeks of low fat diet less than 4.5 is 0.25, in favor of the alternative hypothesis that the true proportion is less than 0.25. Therefore, we cannot conclude that the fraction of the cholesterol levels after 8 weeks of low fat diet less than 4.5 is at most 25%.

```{r setup_diet, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
diet <- read.table("resources/diet.txt", header = TRUE)
diet <- diet %>% 
  mutate(weight_loss = preweight - weight6weeks)
```

# Diet

Dataset peak:  


```{r head_diet, include=TRUE}
head(diet)

```


## a) Informative graphical summary

We will visualise the weight loss and the average weight loss with respect to each diet.


```{r}

box_plot <- diet %>%
  ggplot(aes(x = factor(diet), y = weight_loss)) +
  geom_boxplot() +
  labs(x = "Diet", y = "Weight lost", title = "Weight lost with respect to diet")

median_plot <- diet %>% 
  group_by(diet) %>% 
  summarize(median_weight_loss = median(weight_loss)) %>%
  ggplot(aes(x=factor(diet), y=median_weight_loss)) +
  geom_bar(stat="identity", fill="blue") +
  labs(x="Diet", y ="Median weight loss", title="Median weight loss with respect to diet") 

median_plot + box_plot

```


Based on the visual information of weight loss and medium weight loss we argue that the third diet was the most efficient!


## b) One-way Anova to test whether the the diet has an effect on the weight loss

We assume that the weight loss is significantly different across all types of diet.

```{r}
fit <- aov(weight_loss ~ diet , data = diet)
summary(fit)
```

We observe a very low value for P(F) << 1%, thus we will reject our hypothesis and conclude that at least one of the diets must have an effect over the weight loss.

Given the high mean values for weight loss with respect to diet we conclude that the best diets were: 3, 2 and 1:

```{r}
diet %>% 
  group_by(diet) %>% 
  summarize(median_weight_loss = median(weight_loss)) %>%
  arrange(desc(median_weight_loss))
```



## b) Two-way ANOVA to investigate effect of the diet and gender

```{r}
fit <- aov(weight_loss ~ diet * gender, data = diet)
summary(fit)

```

We observe:

* A very low value P(F) << 1% for diet
  - We conclude diet has an effect on weight loss
* A very high value P(F) >> 85% for gender
  - We conclude gender has no significant effect on weight loss
* A low value P(F) << 5% for the factorized effect of diet & gender on weight loss. 
  - We conclude that there must be a causal effect between the gender and a particular diet.

## e) Preferred ANOVA

The results of the Two-way ANOVA suggest that the effectiveness of a diet is significantly associated with the gender of the participant.

***
# Yield of peas

##A)
Load the MASS package to access the npk dataset
```{r}
library(MASS)
```
Create a vector of plot labels
```{r}
plot_labels <- rep(1:24, each = 3)
```
Randomly permute the plot labels
```{r}
set.seed(123) # for reproducibility
plot_labels <- sample(plot_labels)
```
Assign treatments to each plot based on the randomized plot labels
```{r}
treatments <- rep(c("N", "P", "K"), length.out = 72)
treatments <- treatments[plot_labels]
```
Create a data frame with the randomized plot labels and treatments
```{r}
npk_rand <- data.frame(plot = plot_labels, treatment = treatments)
```
Use reshape() function from the tidyr package to create a wide format
data frame with one row per plot and columns for each treatment
```{r}
library(tidyr)
npk_rand_wide <- reshape(npk_rand, direction = "wide", idvar = "plot", timevar = "treatment")
```
View the randomized treatment assignments
```{r}
npk_rand_wide
```
##B)
Calculate the average yield per block for each treatment
```{r}
library(dplyr)
npk_avg <- npk %>% 
  group_by(block, N) %>% 
  summarize(avg_yield = mean(yield))
```
Make a plot of the average yield per block for each treatment
```{r}
library(ggplot2)
ggplot(npk_avg, aes(x = factor(block), y = avg_yield, fill = factor(N))) +
  geom_col(position = "dodge") +
  labs(x = "Block", y = "Average yield (pounds per plot)", fill = "Nitrogen") +
  scale_fill_discrete(name = "Nitrogen\napplied", labels = c("No", "Yes"))
```
This code creates a data frame npk_avg that contains the average yield per block for each treatment combination (nitrogen applied or not applied) using the dplyr package. The code then uses ggplot2 to create a bar plot of the average yield per block, with separate bars for the two nitrogen treatments, and with blocks on the x-axis. The position = "dodge" argument in geom_col() ensures that the bars for the two nitrogen treatments are displayed side-by-side for each block.

By including the factor block in the analysis, we account for any variability in yield that may be due to differences between blocks (e.g., soil type, sun exposure, etc.). This allows us to focus on the effect of nitrogen specifically, rather than confounding it with block effects. Additionally, the randomized treatment assignments ensure that any differences in yield between the two nitrogen treatments are not due to systematic differences in the placement of nitrogen-treated plots within blocks.



***
***
***
