---
title: "Data Analysis Report"
output:
  pdf_document: default
---

```{r dependencies, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(patchwork)

```

***

# Birthweights
Todo

***

# Cholesterol
TODO

***

# Diet

```{r setup_diet, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
diet <- read.table("resources/diet.txt", header = TRUE)
diet <- diet %>% 
  mutate(weight_loss = preweight - weight6weeks)
```


Dataset peak:  


```{r head_diet, include=TRUE}
head(diet)

```


## a) Informative graphical summary

We will visualise the weight loss and the average weight loss with respect to each diet.


```{r}

box_plot <- diet %>%
  ggplot(aes(x = factor(diet), y = weight_loss)) +
  geom_boxplot() +
  labs(x = "Diet", y = "Weight lost", title = "Weight lost with respect to diet")

median_plot <- diet %>% 
  group_by(diet) %>% 
  summarize(median_weight_loss = median(weight_loss)) %>%
  ggplot(aes(x=factor(diet), y=median_weight_loss)) +
  geom_bar(stat="identity", fill="blue") +
  labs(x="Diet", y ="Median weight loss", title="Median weight loss with respect to diet") 

median_plot + box_plot

```


Based on the visual information of weight loss and medium weight loss we argue that the third diet was the most efficient!


## b) One-way Anova to test whether the the diet has an effect on the weight loss

We assume that the weight loss is significantly different across all types of diet.

```{r}
fit <- aov(weight_loss ~ diet , data = diet)
summary(fit)
```

We observe a very low value for P(F) << 1%, thus we will reject our hypothesis and conclude that at least one of the diets must have an effect over the weight loss.

Given the high mean values for weight loss with respect to diet we conclude that the best diets were: 3, 2 and 1:

```{r}
diet %>% 
  group_by(diet) %>% 
  summarize(median_weight_loss = median(weight_loss)) %>%
  arrange(desc(median_weight_loss))
```



## b) Two-way ANOVA to investigate effect of the diet and gender

```{r}
fit <- aov(weight_loss ~ diet * gender, data = diet)
summary(fit)

```

We observe:

* A very low value P(F) << 1% for diet
  - We conclude diet has an effect on weight loss
* A very high value P(F) >> 85% for gender
  - We conclude gender has no significant effect on weight loss
* A low value P(F) << 5% for the factorized effect of diet & gender on weight loss. 
  - We conclude that there must be a causal effect between the gender and a particular diet.

## e) Preferred ANOVA

The results of the Two-way ANOVA suggest that the effectiveness of a diet is significantly associated with the gender of the participant.

***
# Yield of peas

TODO


***
***
***
